---
title: "CRTrecruit"
author: "Georgia Papadogeorgou"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CRTrecruit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This R package performs correction for recruitment bias in cluster randomized trials where available data are on the recruited units only.

### Why use it:

- Clusters are randomly assigned to treatment or controls. Once cluster-level assignment is perfromed, individuals are recruited in the study.
- Recruitment is generally not blinded to the treatment level of the cluster. This creates post-treatment differences in the treated and control recruited populations. These differences render simple contrasts of outcomes in treated and control recruited groups not causally interpretable.

### What are the situations that CRTrecruit should be used?

- In cases with post-treatment recruitment differences, CRTrecruit can be used to estimate causal effects on interpretable subpopulations.
- These subpopulations correspond to the combination of always- and defier-recruited, the always- and complier-recruited, and the recruited population. When defiers do not exist, such as in the case where treatment can be thought to only lead to increase in the recruitment indicator, the always- and defier-recruited population is the same as the always-recruited population, and we can further estimate effects on the complier-recruited population.
- Our approach relies on assumptions on recruitment ignorability and positivity.

# Using CRTrecruit

## The data set

The data set need to provide information on

- The treatment indicator of the recruited units ```Zobs```
- The outcome of the recruited units ```Yobs```
- The covariate matrix of the recruited units ```Xobs``` where the first column is a column of 1s corresponding to intercepts. These covariates are assumed to satisfy the ignorability (or the non-differential recruitment) assumption.
- Indices of cluster membership for the recruited units ```IDobs```. These indices should be integers starting at 1 and going to the maximum number of clusters.
- The probability of treatment in the cluster assignment ```treat_prop```.

A simulated data set is included in the R package. We can load it using



```{r}
library(CRTrecruit)
data('dta')

# The data only for those who enrolled:
Zobs <- dta$Zobs
Yobs <- dta$Yobs
Xobs <- dta$Xobs
IDobs <- dta$IDobs
treat_prop <- dta$treat_prop
```

and investigate the data structure
 
```{r}
dta$Zobs[1:55]  # The first 55 units
dta$Yobs[1:10]  # The first 10 units
head(dta$Xobs)
dta$IDobs[1:55]  # The first 55 units
```


## The working propensity score

The estimation procedure relies on estimating the working propensity score. We provide code to estimate it using the parametric approach in the manuscript. Specifically:

```{r}

r <- treat_prop / (1 - treat_prop)

# Estimating the propensity score:
est_wps_pars <- EstWPSpars(Xobs = Xobs, Zobs = Zobs, IDobs = IDobs,
                           treat_prop = treat_prop)

# The estimated parameters of the working propensity score:
alphas <- est_wps_pars$est_delta_coefs

# Acquiring the estimates of the working propensity score.
est_delta <- 1 + exp(Xobs %*% alphas)  # The delta function
est_eind <- r * est_delta / (1 + r * est_delta)  # The working PS.
```


## Estimating the causal effects

- Once the working propensity scores have been acquired, we can create the weights that are used for estimating the causal effects in the different populations.
- The weights are calculated as follows:

```{r}
# The weights for the always+defiers, always+compliers, and recruited.
weights <- cbind(alw_def = ((1 - est_eind) / est_eind) * Zobs + (1 - Zobs),
                 alw_com = Zobs + (1 - Zobs) * est_eind / (1 - est_eind),
                 recr = Zobs / est_eind + (1 - Zobs) / (1 - est_eind))
```

### Effects on three interpretable supopulations

- These weights are fed into the ```SimultEstimateCRT``` function for acquiring the causal effect estimates on the three subpopulation, the always- and defier-recruited, the always- and complier-recruited, and the recruited populations.

```{r}
simult_est <- SimultEstimateCRT(Zobs = Zobs, Yobs = Yobs, weights = weights,
                                IDobs = IDobs, Xobs = Xobs,
                                est_wps_pars = est_wps_pars,
                                ps = 'estimated', treat_prop = treat_prop)
```

- The results include the estimates, asymptotic variance, and confidence intervals based on the asymptotic distribution.

```{r}
# The estimates:
simult_est$estimate
# The asymptotic variance:
simult_est$asym_var_est
# The confidence intervals' lower and upper bounds
rbind(lb = simult_est$lb, ub = simult_est$ub)
```


### Effects on the compliers

- It is common to assume monotonicity of the recruitment indicators, implying that recruitment status under treatment can only be equal or higher than recruitment status under control. This assumption rules out the presence of defiers. This means that the effect on the always- and defier-recruited population is simply the effect on the always-recruited population.
- This allows us to also estimate the effect on the complier population.
- To do so, we first need to estimate the proportion of compliers in the group of always- and complier-recruited units. We can do so as follows:

```{r}
p_trt <- mean(Zobs)  # Observed proportion of treatment in the recruited.
pi_trt <- treat_prop  # The theoretical treatment proportion

# The estimated proportion of compliers among always + compliers
est_prop_compl <- 1 - (pi_trt / (1 - pi_trt)) * ((1 - p_trt) / p_trt)
```

- Then, we can use the ```ComplierEffect``` function to acquire the causal effect on the complier group.

```{r}
compl <- ComplierEffect(prop_compl = est_prop_compl,
                        effect_estimates = simult_est$estimate[1:2],
                        asym_var_est = simult_est$asym_var_est[1 : 2, 1 : 2])
compl
```


## Inference using the bootstrap

- Bootstrap based confidence intervals can also be acquired.
- We resample clusters, either completely randomly by setting ```fix_trt_num = FALSE``` or by keeping the number of treated clusters fixed ```fix_trt_num = TRUE``` (default).
- If we assume that there are no defiers (```defiers=FALSE```) the bootstrap will also calculate bootstrap estimates for the effect on the compliers.

```{r}
boots <- BootVar(B = 500, Xobs = Xobs, Zobs = Zobs, Yobs = Yobs, IDobs = IDobs,
                 treat_prop = treat_prop, fix_trt_num = TRUE,
                 prop_compl = est_prop_compl, defiers = FALSE, verbose = TRUE)
```

- We create confidence intervals using the standard deviation of the bootstrap samples and using the normal approximation.
- Quantiles could be considered alternatively.

```{r}
boot_sd <- apply(boots$est_boot, 2, sd)[1 : 4]
boot_sd <- boot_sd[c(1, 3, 2, 4)]  # Re-ordering.
effects <- c(simult_est$estimate, compliers = as.numeric(compl[1]))

effects
boot_sd

rbind(lb = effects - 1.96 * boot_sd, ub = effects + 1.96 * boot_sd)
```


